Reliable error estimation for Sobol' indices

### 2016-08-30 article/template.tex
last check

\newcommand{\fudge}{\fC}
\newcommand{\cc}{\mathcal{C}}
\newcommand{\dtf}{\textit{\doubletilde{f}}}
\newcommand{\cube}{[0,1]^d}
\newcommand{\rf}{\mathring{f}}
\newcommand{\rnu}{\mathring{\nu}}
\newcommand{\natm}{\naturals_{0,m}}
\newcommand{\wcS}{\widecheck{S}}
\newcommand{\tol}{\text{tol}}
\newcommand{\e}{\text{e}}
%\newcommand{\bvec}[1]{\boldsymbol{#1}}
\newcommand{\vect}[1]{\boldsymbol{\mathbf{#1}}}
\newcommand{\vx}{\vect{x}}
\newcommand{\vI}{\vect{I}}
\newcommand{\vk}{\vect{k}}
\newcommand{\vw}{\vect{w}}
\newcommand{\vz}{\vect{z}}
\newcommand{\vt}{\vect{t}}
\newcommand{\dif}{\mathsf{d}}
\newcommand{\hf}{\hat{f}}
\newcommand{\hS}{\widehat{S}}
\newcommand{\tS}{\widetilde{S}}
\newcommand{\tf}{\tilde{f}}
\newcommand{\fC}{\mathfrak{C}}
\newcommand{\homega}{\widehat{\omega}}
\newcommand{\wcomega}{\mathring{\omega}}
\newcommand{\vzero}{\vect{0}}
\newcommand{\integers}{\mathbb{Z}}
\newcommand{\naturals}{\mathbb{N}}
\newcommand{\ip}[3][{}]{\ensuremath{\left \langle #2, #3 \right \rangle_{#1}}}
\newcommand\iid{\stackrel{{\rm iid}}{\sim}}

@@

The problem of interest is the evaluation of first-order and total effect Sobol' indices. In our framework, we are only interested in single input indices, $|u|=1$. To this effect we restrain $u \in \mathcal{D}$. The computation of the normalized indices is performed based on the following integral formulas for their numerators:
\begin{align}
\label{first.order}
\underline{\tau}_u^2  &= \int_{[0,1)^{2d-1}} \left(f(\vx)-
f(\vx_u:{\vx'}_{-u})\right)f(\vx')d\vx d{\vx'_{-u}}, \\
\label{total.effect}
\ov{$\tau$}_u^2 &= \frac{1}{2}\int_{[0,1)^{d+1}}(f(\vx')-f(\vx_u:{\vx'}_{-u}))^2 d\vx_u d{\vx'}, \ u \in \mathcal{D},
\end{align}
while variance and mean of $f$ are evaluated as:
\begin{gather}
\begin{aligned}
\sigma^2 &= \int_{[0,1)^{d}} f(\vx)^2d{\vx} - \mu^2, \\
 \mu &= \int_{[0,1)^{d}} f(\vx) d{\vx},
\label{eq.sigma.mu}
\end{aligned}
\end{gather}
Usually the complexity of $f$ causes the computation of integrals \eqref{first.order}, (\ref{total.effect}) and (\ref{eq.sigma.mu}) to be intractable. In such cases, one can instead estimate these quantities.

@@

Hence, in \cite[Sec. 4.2]{HicJim} it is proved that for any $f\in\cc$,
\begin{equation}
\absolute{I(f) - \frac{1}{b^m}\sum_{\vx\in\mathcal{P}_m}f(\vx)}
\le \overbrace{\frac{\tS_{\ell,m}(f)\homega(m) \wcomega(m-\ell)}{1 - \homega(m-\ell) \wcomega(m-\ell)}}^{\epsilon_{I(f)}=}, \label{errbd}
\end{equation}
where one may increase $m$ until the error bound $\epsilon_{I(f)}$ is small enough.

@@

If Variant $B$ is selected the cost of our algorithm equals $2 \times 2^{m^{\star}}$. This cost corresponds to the one of the replication procedure claimed in the introduction, where $2n=2\times 2^{m^{\star}}$ independent of $d$.

@@

Assume that the number of small first-order indices is known and equals $\gamma$. We denote by $u_1,\dots,u_{\gamma}$ the indices of the corresponding inputs and $\Gamma = \{1,\dots,\gamma\}$. The cost of Variant $A$ including ``Correlation 2" writes: 
\begin{equation}
\sum \limits_{j \in \Gamma} 2^{m''_{u_j}} + \sum \limits_{j \in \Gamma} 2^{m'_{u_j}} + \sum \limits_{j \in \mathcal{D}\setminus{\Gamma}} 2^{m_{u_j}} + 2 \times 2^{m^\star},
\label{cost.improvement}
\end{equation}
where $m'_{u_j}\leq m^{\star}$, $\ m_{u_j} \leq m^{\star}$, and: 
\begin{itemize}
\item[$\bullet$] for $j\in \Gamma$, $2^{m''_{u_j}}$ is the number of evaluations \\ $f\left(\vz_{i,u_j}:\vx_{i,-u_j}\right)$ to estimate $\underline{S}_{u_j}$,
\item[$\bullet$] for $j\in \Gamma$, $2^{m'_{u_j}}$ is the number of evaluations \\ $f\left(\vx_{i,u_j}:{\vx'}_{i,-u_j}\right)$ to estimate both $\underline{S}_{u_j}$ and  $\overline{S}_{u_j}$,
\item[$\bullet$] likewise, for $j\in \mathcal{D}\setminus{\Gamma}$, $ 2^{m_{u_j}}$ is the number of evaluations $f\left(\vx_{i,u_j}:{\vx'}_{i,-u_j}\right)$ to estimate both $\underline{S}_{u_j}$ and  $\overline{S}_{u_j}$,
\item[$\bullet$] $2 \times 2^{m^{\star}}$ is the number of evaluations $f\left(\vx_{i}\right)$ and $f\left({\vx'}_{i}\right)$ used in the estimation of each first-order and total effect index.
\end{itemize}
Recall that the cost of Variant $A$ without  ``Correlation 2" writes:
\begin{equation}
\sum \limits_{j \in \Gamma} 2^{m_{u_j}} + \sum \limits_{j \in \mathcal{D}\setminus{\Gamma}} 2^{m_{u_j}} + 2 \times 2^{m^{\star}}, \qquad m_{u_j} \leq m^{\star}.
\label{cost.non.improvement}
\end{equation}
The difference between costs (\ref{cost.improvement}) and (\ref{cost.non.improvement}) equals:
\begin{equation}
 \sum \limits_{j \in \Gamma} 2^{m_{u_j}} \left(2^{m''_{u_j}-m_{u_j}}  + 2^{m'_{u_j}-m_{u_j}} - 1 \right) = \sum \limits_{j \in \Gamma} c_j.
\label{cost.comparison}
\end{equation}
Hence, the sign of this difference indicates whether or not using ``Correlation 2" brings an improvement to Variant $A$. We distinguish two cases :
\begin{itemize}
\item[1)] for $j \in \Gamma$, the total effect index $\ov{$S$}_{u_j}$ requires as much or more evaluations than the first-order index $\underline{S}_{u_j}$. Since the total effect estimator is the same, as a consequence we have $m'_{u_j}=m_{u_j}$ and $c_j >0$.
\item[2)] for $j \in \Gamma$, the total effect index $\ov{$S$}_{u_j}$ requires less evaluations than the first-order index $\underline{S}_{u_j}$. In this case, if both $m''_{u_j} < m_{u_j}$ and $m'_{u_j} < m_{u_j}$ then $c_j \leq 0$.  
\end{itemize}
Overall we expect to observe case $2)$ more often than case $1)$. Indeed, the numerator of $\underline{S}_u$ requires to estimate $2d-1$ dimensional integrals against only $d+1$ dimensional integrals for the numerator of $\ov{$S$}_u$. Hence, it seems reasonable to expect that it will take less points to estimate $\ov{$S$}_u$ than $\underline{S}_u$.

Furthermore, in case $2)$, we expect the two conditions $m''_{u_j} < m_{u_j}$ and $m'_{u_j} < m_{u_j}$ to always hold as ``Correlation 2" is shown 
to perform better for small first-order indices. To support this discussion, illustrations of the use of ``Correlation 2" are presented in Section \ref{sec:5}.

In practice, one does not know which are the small Sobol' indices. To overcome this issue, we propose the following alternative for Variant $A$. If at the end of the first iteration, the estimate $\widehat{\underline{S}}_u$, $u \in \mathcal{D}$, is smaller than a threshold, then at the next iteration estimator (\ref{first.order.est}) is switched for ``Correlation 2" and a third Sobol' sequence $\mathcal{P''}_{m}=\{\vz_i\}_{i=0}^{2^m-1}$ is constructed to obtain the corresponding evaluations $f({\vz}_{i,u}:{\vx}_{i,-u})$.

@@

% Non-BibTeX users please use
\begin{thebibliography}{99}
%
% and use \bibitem to create references. Consult the Instructions
% for authors for reference list style.
%
\bibitem{Gilquin.rec}
Gilquin, L., Arnaud, E., Monod, H., Prieur, C.: Recursive estimation procedure of Sobol' indices based on replicated designs, preprint available at \url{https://hal.inria.fr/hal-01291769}, (2016).
\bibitem{GJAHMP}
Gilquin, L., Jim\'enez Rugama, Ll.A., Arnaud, E., Hickernell, F.J., Monod, H., Prieur, C.: Iterative construction of replicated designs based on Sobol' sequences, preprint available at \url{https://hal.inria.fr/hal-01349444} (2016)
\bibitem{HicJim}
Hickernell, F.J., Jim\'enez Rugama, Ll.A.: Reliable Adaptative Cubature Using Digital Sequences: Monte Carlo and Quasi-Monte Carlo Methods, vol. 163, 367-383 (2016)
\bibitem{Hoeffding}
Hoeffding, W.F.: A class of statistics with asymptotically normal distribution, Ann. Math. Stat. \textbf{19}(3), 293-325 (1948)
%\bibitem{Janon}
%Janon, A., Klein, T., Lagnoux A., Nodet, M., Prieur C.: Asymptotic normality and efficiency of two {S}obol' index estimators, ESAIM Probab. Stat. \textbf{18}, 342-364 (2014)
\bibitem{Lemieuxbook}
Lemieux,C.: Monte Carlo and Quasi-Monte Carlo Sampling, Springer, New-York (2009)
\bibitem{Mara}
Mara, T.A., Rakoto Joseph, 0.: Comparison of some efficient methods to evaluate the main effect of computer model factors, J. Statist. Comput. Simulation \textbf{78}(2), 167-178 (2008)
\bibitem{McKay}
McKay, M.D.: Evaluating prediction uncertainty, Los Alamos National Laboratory Report NUREG/CR- 6311, LA-12915-MS. (1995)
\bibitem{Niederreiter} 
Niederreiter, H.: Random Number Generation and Quasi-Monte Carlo Methods: CBMS-NSF Regional Conference Series in Applied Math., vol. 63, SIAM, Philadelphia (1992)
\bibitem{Owen}
Owen, A.B.: Better estimation of small Sobol' sensitivity indices, ACM Trans. Model. Comput. Simul. \textbf{23}(2), :11 (2013)
\bibitem{Saltelli}
Saltelli, A.: Making best use of models evaluations to compute sensitivity indices, Comput. Phys. Commun. \textbf{145}(2), 280-297 (2002)
\bibitem{Sobol'seq}
Sobol', I.M.: On the distribution of points in a cube and the approximate evaluation of integrals, USSR Comput. Math. Math. Phys. \textbf{7}(4), 86-112 (1967)
\bibitem{Sobol'}
Sobol', I.M.: Sensitivity indices for nonlinear mathematical models, Mathematical Modeling and Computational Experiment \textbf{1}, 407-414 (1993)
\bibitem{Tissot}
Tissot, J.Y., Prieur, C.: A randomized Orthogonal Array-based procedure for the estimation of first- and second-order Sobol' indices, J. Statist. Comput. Simulation \textbf{85}(7), 1358-1381 (2015)
\bibitem{Tong}
Tong C.: Self-validated variance-based methods for sensitivity analysis of model outputs, Reliab. Eng. Syst. Saf. \textbf{95}(3), 301-309 (2010)

>>>>

\newcommand{\fudge}{\fC}
\newcommand{\cc}{\mathcal{C}}
\newcommand{\dtf}{\textit{\doubletilde{f}}}
\newcommand{\cube}{[0,1]^d}
\newcommand{\rf}{\mathring{f}}
\newcommand{\rnu}{\mathring{\nu}}
\newcommand{\natm}{\naturals_{0,m}}
\newcommand{\wcS}{\widecheck{S}}
\newcommand{\tol}{\text{tol}}
\newcommand{\e}{\text{e}}
%\newcommand{\bvec}[1]{\boldsymbol{#1}}
\newcommand{\vect}[1]{\boldsymbol{\mathbf{#1}}}
\newcommand{\vx}{\vect{x}}
\newcommand{\vI}{\vect{I}}
\newcommand{\vk}{\vect{k}}
\newcommand{\vw}{\vect{w}}
\newcommand{\vz}{\vect{z}}
\newcommand{\vt}{\vect{t}}
\newcommand{\dif}{\mathsf{d}}
\newcommand{\hf}{\hat{f}}
\newcommand{\hS}{\widehat{S}}
\newcommand{\tS}{\widetilde{S}}
\newcommand{\tf}{\widetilde{f}}
\newcommand{\fC}{\mathfrak{C}}
\newcommand{\homega}{\widehat{\omega}}
\newcommand{\wcomega}{\mathring{\omega}}
\newcommand{\vzero}{\vect{0}}
\newcommand{\integers}{\mathbb{Z}}
\newcommand{\naturals}{\mathbb{N}}
\newcommand{\ip}[3][{}]{\ensuremath{\left \langle #2, #3 \right \rangle_{#1}}}
\newcommand\iid{\stackrel{{\rm iid}}{\sim}}

@@

The problem of interest is the evaluation of first-order and total effect Sobol' indices. In our framework, we are only interested in single input indices, namely $|u|=1$. The computation of the normalized indices is performed based on the following integral formulas for their numerators:
\begin{align}
\label{first.order}
\underline{\tau}_u^2  &= \int_{[0,1)^{2d-1}} \left(f(\vx)-
f(\vx_u:{\vx'}_{-u})\right)f(\vx')d\vx d{\vx'_{-u}}, \\
\label{total.effect}
\ov{$\tau$}_u^2 &= \frac{1}{2}\int_{[0,1)^{d+1}}(f(\vx')-f(\vx_u:{\vx'}_{-u}))^2 d\vx_u d{\vx'}, \ u \in \mathcal{D},
\end{align}
while variance and mean of $f$ are evaluated as:
\begin{gather}
\begin{aligned}
\sigma^2 &= \int_{[0,1)^{d}} f(\vx)^2d{\vx} - \mu^2, \\
 \mu &= \int_{[0,1)^{d}} f(\vx) d{\vx},
\label{eq.sigma.mu}
\end{aligned}
\end{gather}
Usually the complexity of $f$ causes the computation of integrals \eqref{first.order}, (\ref{total.effect}) and (\ref{eq.sigma.mu}) to be intractable. In such cases, one can instead estimate these quantities.

@@

Hence, in \cite[Sec. 4.2]{HicJim} it is proved that for any $f\in\cc$,
\begin{equation}
\absolute{I(f) - \frac{1}{b^m}\sum_{\vx\in\mathcal{P}_m}f(\vx)}
\le \frac{\tS_{\ell,m}(f)\homega(m) \wcomega(m-\ell)}{1 - \homega(m-\ell) \wcomega(m-\ell)}=\epsilon_{I(f)}, \label{errbd}
\end{equation}
where one may increase $m$ until the error bound $\epsilon_{I(f)}$ is small enough.

@@

If Variant $B$ is selected the cost of our algorithm equals $2 \times 2^{m^{\star}}$. This cost corresponds to the one of the replication procedure introduced in Section \ref{sec:2.2}, where $2n=2\times 2^{m^{\star}}$ independent of $d$.

@@

Assume that the number of small first-order indices is known and equals $\gamma$. We denote by $u_1,\dots,u_{\gamma}$ the indices of the corresponding inputs and $\Gamma = \{1,\dots,\gamma\}$. The cost of Variant $A$ including ``Correlation 2" writes: 
\begin{equation}
\sum \limits_{j \in \Gamma} 2^{m''_{u_j}} + \sum \limits_{j \in \Gamma} 2^{m'_{u_j}} + \sum \limits_{j \in \mathcal{D}\setminus{\Gamma}} 2^{m_{u_j}} + 2 \times 2^{m^\star},
\label{cost.improvement}
\end{equation}
where: %$m'_{u_j}\leq m^{\star}$, $\ m_{u_j} \leq m^{\star}$, and: 
\begin{itemize}
\item[$\bullet$] for $j\in \Gamma$, $2^{m''_{u_j}}$ is the number of evaluations \\ $f\left(\vz_{i,u_j}:\vx_{i,-u_j}\right)$ to estimate $\underline{S}_{u_j}$,
\item[$\bullet$] for $j\in \Gamma$, $2^{m'_{u_j}}$ is the number of evaluations \\ $f\left(\vx_{i,u_j}:{\vx'}_{i,-u_j}\right)$ to estimate both $\underline{S}_{u_j}$ and  $\overline{S}_{u_j}$,
\item[$\bullet$] likewise, for $j\in \mathcal{D}\setminus{\Gamma}$, $ 2^{m_{u_j}}$ is the number of evaluations $f\left(\vx_{i,u_j}:{\vx'}_{i,-u_j}\right)$ to estimate both $\underline{S}_{u_j}$ and  $\overline{S}_{u_j}$,
\item[$\bullet$] $2 \times 2^{m^{\star}}$ is the number of evaluations $f\left(\vx_{i}\right)$ and $f\left({\vx'}_{i}\right)$ used in the estimation of each first-order and total effect index.
\end{itemize}
Recall that the cost of Variant $A$ without  ``Correlation 2" writes:
\begin{equation}
\sum \limits_{j \in \Gamma} 2^{m_{u_j}} + \sum \limits_{j \in \mathcal{D}\setminus{\Gamma}} 2^{m_{u_j}} + 2 \times 2^{m^{\star}}. 
%\qquad m_{u_j} \leq m^{\star}.
\label{cost.non.improvement}
\end{equation}
The difference (\ref{cost.improvement}) $-$ (\ref{cost.non.improvement}) equals:
\begin{equation}
 \sum \limits_{j \in \Gamma} 2^{m_{u_j}} \left(2^{m''_{u_j}-m_{u_j}}  + 2^{m'_{u_j}-m_{u_j}} - 1 \right) = \sum \limits_{j \in \Gamma} c_j.
\label{cost.comparison}
\end{equation}
Hence, the sign of this difference indicates whether or not using ``Correlation 2" brings an improvement to Variant $A$. We distinguish two cases :
\begin{itemize}
\item[1)] for $j \in \Gamma$, the total effect index $\ov{$S$}_{u_j}$ requires as much or more evaluations than the first-order index $\underline{S}_{u_j}$. Since the total effect estimator is the same, as a consequence we have $m'_{u_j}=m_{u_j}$ and $c_j >0$.
\item[2)] for $j \in \Gamma$, the total effect index $\ov{$S$}_{u_j}$ requires less evaluations than the first-order index $\underline{S}_{u_j}$. In this case, if both $m''_{u_j} < m_{u_j}$ and $m'_{u_j} < m_{u_j}$ then $c_j \leq 0$.  
\end{itemize}
Overall we expect to observe case $2)$ more often than case $1)$. Indeed, the numerator of $\underline{S}_u$ requires to estimate $2d-1$ dimensional integrals against only $d+1$ dimensional integrals for the numerator of $\ov{$S$}_u$. Hence, it seems reasonable to expect that it will take less points to estimate $\ov{$S$}_u$ than $\underline{S}_u$.

Furthermore, in case $2)$, we expect the two conditions $m''_{u_j} < m_{u_j}$ and $m'_{u_j} < m_{u_j}$ to always hold as ``Correlation 2" is shown in \cite{Owen} to perform better for small first-order indices. To support this discussion, illustrations of the use of ``Correlation 2" are presented in Section \ref{sec:5}.

In practice, one does not know which are the small Sobol' indices ($u_1,\dots,u_\gamma)$. To overcome this issue, we propose the following alternative for Variant $A$. If at the end of the first iteration, the estimate $\widehat{\underline{S}}_u$, $u \in \mathcal{D}$, is smaller than a threshold (here $0.1$), then at the next iteration estimator (\ref{first.order.est}) is switched for ``Correlation 2" and a third Sobol' sequence $\mathcal{P''}_{m}=\{\vz_i\}_{i=0}^{2^m-1}$ is constructed to obtain the corresponding evaluations $f({\vz}_{i,u}:{\vx}_{i,-u})$.

@@

% Non-BibTeX users please use
\begin{thebibliography}{99}
%
% and use \bibitem to create references. Consult the Instructions
% for authors for reference list style.
%
\bibitem{Gilquin.rec}
Gilquin, L., Arnaud, E., Monod, H., Prieur, C.: Recursive estimation procedure of Sobol' indices based on replicated designs, preprint available at \url{https://hal.inria.fr/hal-01291769}, (2016).
\bibitem{GJAHMP}
Gilquin, L., Jim\'enez Rugama, Ll.A., Arnaud, E., Hickernell, F.J., Monod, H., Prieur, C.: Iterative construction of replicated designs based on Sobol' sequences, preprint available at \url{https://hal.inria.fr/hal-01349444} (2016)
\bibitem{HicJim}
Hickernell, F.J., Jim\'enez Rugama, Ll.A.: Reliable Adaptative Cubature Using Digital Sequences: Monte Carlo and Quasi-Monte Carlo Methods, vol. 163, 367-383 (2016)
\bibitem{Hoeffding}
Hoeffding, W.F.: A class of statistics with asymptotically normal distribution, Ann. Math. Stat. \textbf{19}(3), 293-325 (1948)
\bibitem{Janon}
Janon, A., Klein, T., Lagnoux A., Nodet, M., Prieur C.: Asymptotic normality and efficiency of two {S}obol' index estimators, ESAIM Probab. Stat. \textbf{18}, 342-364 (2014)
\bibitem{Lemieuxbook}
Lemieux,C.: Monte Carlo and Quasi-Monte Carlo Sampling, Springer, New-York (2009)
\bibitem{Mara}
Mara, T.A., Rakoto Joseph, 0.: Comparison of some efficient methods to evaluate the main effect of computer model factors, J. Statist. Comput. Simulation \textbf{78}(2), 167-178 (2008)
\bibitem{McKay}
McKay, M.D.: Evaluating prediction uncertainty, Los Alamos National Laboratory Report NUREG/CR- 6311, LA-12915-MS. (1995)
\bibitem{Niederreiter} 
Niederreiter, H.: Random Number Generation and Quasi-Monte Carlo Methods: CBMS-NSF Regional Conference Series in Applied Math., vol. 63, SIAM, Philadelphia (1992)
\bibitem{Owen}
Owen, A.B.: Better estimation of small Sobol' sensitivity indices, ACM Trans. Model. Comput. Simul. \textbf{23}(2), :11 (2013)
\bibitem{Saltelli}
Saltelli, A.: Making best use of models evaluations to compute sensitivity indices, Comput. Phys. Commun. \textbf{145}(2), 280-297 (2002)
\bibitem{Sobol'seq}
Sobol', I.M.: On the distribution of points in a cube and the approximate evaluation of integrals, USSR Comput. Math. Math. Phys. \textbf{7}(4), 86-112 (1967)
\bibitem{Sobol'}
Sobol', I.M.: Sensitivity indices for nonlinear mathematical models, Mathematical Modeling and Computational Experiment \textbf{1}, 407-414 (1993)
\bibitem{Tissot}
Tissot, J.Y., Prieur, C.: A randomized Orthogonal Array-based procedure for the estimation of first- and second-order Sobol' indices, J. Statist. Comput. Simulation \textbf{85}(7), 1358-1381 (2015)
\bibitem{Tong}
Tong C.: Self-validated variance-based methods for sensitivity analysis of model outputs, Reliab. Eng. Syst. Saf. \textbf{95}(3), 301-309 (2010)
